\documentclass[12pt]{article}
\usepackage[top=1in, bottom=1in, left=1in, right=1in]{geometry}

\title{NE 155, Class 3, S14 \\
Types of Equations in the Engineering Fields}
\author{Rachel Slaybaugh}
\date{January 27, 2014}

\usepackage{setspace}
\onehalfspacing

\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
\usepackage{amsthm}
\usepackage{epsfig}
\usepackage{times}
\renewcommand{\ttdefault}{cmtt}
\usepackage{amsmath}
\usepackage{graphicx} % for graphics files

% Draw figures yourself
\usepackage{tikz} 

% The float package HAS to load before hyperref
\usepackage{float} % for psuedocode formatting
\usepackage{xspace}

% from Denovo Methods Manual
\usepackage{mathrsfs}
\usepackage[mathcal]{euscript}
\usepackage{color}
\usepackage{array}

\usepackage[pdftex]{hyperref}

\newcommand{\nth}{n\ensuremath{^{\text{th}}} }
\newcommand{\ve}[1]{\ensuremath{\mathbf{#1}}}
\newcommand{\Macro}{\ensuremath{\Sigma}}


\begin{document}
\begin{center}
{\bf NE 155, Class 3, S14 \\
Types of Equations in the Engineering Fields \\ January 27, 2014}
\end{center}

\setlength{\unitlength}{1in}
\begin{picture}(6,.1) 
\put(0,0) {\line(1,0){6.25}}         
\end{picture}


\noindent \textbf{Introduction}

In science and engineering in general, and nuclear engineering and reactor analysis in specific, we encounter a wide range of mathematical physics equations. In today's lecture we will introduce some of them.

\begin{itemize}
\item Ordinary differential equations (ODEs)
\item Partial differential equations (PDEs)
  \begin{itemize}
  \item Elliptic PDEs
  \item Parabolic PDEs
  \item Hyperbolic PDEs
  \end{itemize}
\item Integro-differential equations
\item Integral equations
\end{itemize}

\section{ODEs}

The most general form of an \nth order linear ordinary differential eqn.\ is
%
\begin{equation}
a_{n+1}(x)y^{(n)} + a_{n}(x)y^{(n-1)} + \cdots + a_{2}(x)y^{(1)} + a_{1}(x)y + a_0 = f(x) \nonumber
\end{equation}
%
\noindent where
\begin{itemize}
\item $a_n$ are coefficients
\item $y^{(n)}$ is the \nth derivative of $y$.
\end{itemize}

Boundary conditions:
\begin{enumerate}
\item if y and its derivatives are given at one end of the domain/interval (time 0 if there's time or x 0 if there's only space, etc.): Initial Value Problem (IVP)
\item if y and/or its derivatives are given at \underline{each} end of the interval: Boundary Value Problem (BVP)
\end{enumerate}

\noindent \textbf{Linear 1st order ODE's}

\noindent \underline{Reminders}
\begin{itemize}
\item Linear means each term is either a constant or the product of a constant and the first power of a single variable (in this case y).
\item the highest derivative is to the first power.
\end{itemize}

%-------------------------------------------------------------
\noindent \underline{1st order ODE Example}:
\begin{equation}
y' + 3y = sin(x) \nonumber
\end{equation}
%
\begin{itemize}
\item IVP if boundary conditions are y(0) = 1; y'(0) = 2
\item BVP if boundary conditions are y(0)=-1, y(1) = 3 on the domain [0,1]
\end{itemize}
%
In this case the general solution is obtained through the use of an integrating factor.

\vspace*{1em}
\noindent \underline{1st order ODE Example}:

Point Kinetics analysis of a nuclear reactor is an IVP, linear 1st order ODE.
%
\begin{align}
\frac{dn(t)}{dt} &= \frac{\rho(t) - \beta}{l^*}n(t) + \sum_{i=1}^{N} \lambda_i C_i(t) \nonumber \\
%
\frac{dC_i(t)}{dt} &= \frac{\beta_i}{l^*}n(t) - \lambda_i C_i(t) \qquad i=1,\dots,N \nonumber
\end{align}
%
Where (we'll talk  more about what these terms mean later)
%
\begin{itemize}
\item n = \# neutrons / s
\item $\beta$ = fraction of delayed neutrons
\item $\lambda_i$ = effective decay constant of the ith precursor
\item $C_i(t)$ = delayed neutron concentration of the ith precursor
\item $l^*$ = mean neutron lifetime
\item $\rho = \frac{k-1}{k}$ = reactivity
\end{itemize}
%
BCs: $n(t=0) = n_0$ and $C_i(t=0) = C_{i,0}$ for $i=1,\dots,N$.

%-------------------------------------------------------------
\vspace*{1em}
\noindent \underline{1st order ODE Example}:

The number of atoms in during radioactive decay (assuming decay only here) is described by the Bateman equation, which is a linear, 1st order ODE that is in a IVP:
%
\begin{align}
\frac{dN_1(t)}{dt} &= -\lambda_1 N_1(t) \nonumber \\
\frac{dN_i(t)}{dt} &= -\lambda_i N_i(t) + \lambda_{i-1}N_{i-1}(t) + \alpha_i F_i \qquad 1 < k < I \nonumber\\
\frac{dN_I(t)}{dt} &= -\lambda_{I-1} N_{I-1}(t) \nonumber \\
\text{BC: }& N_i(t=0) = N_{i,0}\nonumber
\end{align}
%
note: isotope $i$ decays into $i+1$. This can be adapted for decay branches, and becomes more complicated if we have neutrons that transmute isotopes. 

%-------------------------------------------------------------
\vspace*{1em}
\noindent \underline{2nd order ODE Example}:

\begin{align}
-\frac{d}{dx}p(x) \frac{d}{dx}\phi(x) &+ q(x)\phi(x) = S(x) \nonumber \\
\text{defined for }& \alpha \le x \le \beta \nonumber \\
\text{BC: }& \frac{d\phi}{dx} + \gamma \phi = \sigma \qquad \text{at } x=\alpha \text{ and } x = \beta \text{ (BVP)}\nonumber
\end{align}
%
This has
\begin{itemize}
\item Neumann BCs if $\gamma = 0$ (specifies the values that the derivative of a solution is to take on the boundary of the domain)
\item Dirichet BCs: if $\gamma \rightarrow \infty, \Macro / \gamma = $constant then the BD reduces to $\phi = $constant (specifies the values that a solution needs to take on the boundary of the domain)
\item Mixed BCs if $\gamma \ne 0$ (different boundary conditions are used on different parts of the boundary of the domain of the equation.)
\item If $S(x)$ is nonzero at least somewhere over the physical range, a unique solution exists.
\end{itemize}

%-------------------------------------------------------
\vspace*{1 em} \underline{Reminder: eigenpairs}

We can formulate systems of equations as matrix-vector systems that look like \ve{A}$x = \lambda x$. 
\begin{itemize}
\item An eigenvector is a non-zero vector $x$ that, when multiplied by the matrix \ve{A}, yields a constant multiple of $x$. 
\item The constant multiple, $\lambda$, is the eigenvalue corresponding to the eigenvector.
\item There can be (and usually are) more than one eigenvalue, and more than one eigenvector. 
%\item Several eigenvectors can have the same eigenvalue
\item Sometime the equation can be reformulated as as $y\ve{A} = \alpha y$ (the left eigenvector). In nuclear, we're used to seeing the right eigenvector formulation.
\end{itemize}
%-------------------------------------------------------------

\vspace*{1em}
\noindent \underline{2nd order ODE Example}:

The homogeneous equation [If  $\phi(x)$  is a solution, so is  $c \phi(x)$, where $c$ is an arbitrary (non-zero) constant. Note that in order for this condition to hold, each term in a linear differential equation of the dependent variable $y$ must contain $y$ or any derivative of $y$; a constant term breaks homogeneity]:
%
\begin{align}
-\frac{d}{dx}p(x) \frac{d}{dx}\phi(x) &+ q(x)\phi(x) = \lambda f(x) \phi(x) \nonumber \\
\text{where }p(x) > 0, f(x) \geq 0, \text{ and }& \lambda = \text{ eigenvalue, and}\nonumber \\
\frac{d\phi}{dx} + \gamma_L \phi &= \Macro \qquad \text{at }x=\alpha \nonumber \\
\frac{d\phi}{dx} + \gamma_R \phi &= \Macro \qquad \text{at }x=\beta \nonumber
\end{align}
%
with $\gamma_L \geq 0$ and $\gamma_R \geq 0$ this is known as the \underline{Sturm-Liouville} eigenvalue problem. If $\gamma_L = \gamma_R = 0$, the BCs become $\phi(\alpha) = \phi(\beta) = 0$.

The solution has an infinite number of eigenfunctions, $\phi_i(x)$ with corresponding \textbf{REAL} and \textbf{DISTINCT} eigenvalues, $\lambda_i$. 

If we number them in sequence: $\lambda_0 < \lambda_1 < \lambda_2 < \dots < \lambda_{\infty}$, then $\lambda_0$ is the lowest eigenvalue and $\phi_0(x)$ is the fundamental eigenmode. 

If $q(x) \geq 0$, then all $\lambda_i$ are positive.

%-------------------------------------------------------------
\vspace*{1em}
\noindent \underline{2nd order ODE Example}:

1-D, 1-group, time-independent neutron diffusion equation:
%
\begin{align}
-\frac{d}{dx}D(x)\frac{d}{dx}\phi(x) + \Macro_a(x)\phi(x) &= S(x) \qquad \text{Fixed Source} \nonumber \\
-\frac{d}{dx}D(x)\frac{d}{dx}\phi(x) + \Macro_a(x)\phi(x) &= \frac{1}{k} \nu \Macro_f(x) \phi(x)\qquad \text{Fission / Eigenvalue} \nonumber
\end{align}
%
BCs: (BVP) vacuum, $\phi(\pm a) = 0$

%-------------------------------------------------------------
\section{PDEs}

A partial differential equation is an equation containing an unknown function of two or more variables and its derivatives with respect to those variables. 

If the PDE is linear in u and all derivatives of u, then we say that the PDE is linear.
%
\begin{equation}
A\frac{\partial^2 u}{\partial x^2} + B\frac{\partial^2 u}{\partial x \partial  y} + C\frac{\partial^2 u}{\partial y^2} + D\frac{\partial u}{\partial x} + E\frac{\partial u}{\partial y} + Fu = G \nonumber
\end{equation}
%
This equation is a 2nd order PDE in two variables. It is linear if $A$ through $G$ depend on $x$ or $y$, but not on $U$.

%-------------------------------------------------------------
\vspace*{1em}
\noindent \underline{Classification}:

\begin{itemize}
\item Elliptic if $B^2 - 4 AC < 0$
\item Hyperbolic if $B^2 - 4 AC > 0$
  \begin{itemize}
  \item if $U$ and its first $x$ derivative are arbitrarily specified initial data on the initial line $x = boundary$ (with sufficient smoothness properties), then there exists a solution for all of $x$.
  \item The solutions of hyperbolic equations are ``wave-like." If a disturbance is made in the initial data of a hyperbolic differential equation, then not every point of space feels the disturbance at once.
  \item Relative to a fixed time coordinate, disturbances have a finite propagation speed. They travel along the characteristics of the equation.
  \end{itemize}
\item Parabolic if $B^2 - 4 AC = 0$\\ A perturbation of the initial (or boundary) data of an elliptic or parabolic equation is felt at once by essentially all points in the domain.
\end{itemize}

\noindent The reason we care based on the TE
\begin{itemize}
\item In a void, the transport equation is like a hyperbolic wave equation. 
\item For highly-scattering regions where $\Macro_{s}$ is close to $\Macro$, the equation becomes elliptic for the steady-state case. 
\item If the scattering is forward-peaked then the equation is parabolic.
\end{itemize}

%-------------------------------------------------------------
\vspace*{1em}
\noindent \underline{Elliptic example}:

1-group, 2-D, steady-state neutron diffusion equation
%
\begin{equation}
\frac{\partial}{\partial x}D\frac{\partial}{\partial x}\phi + \frac{\partial}{\partial y}D\frac{\partial}{\partial y}\phi + (\frac{1}{k} \nu \Macro_f(x) - \Macro_a(x)) \phi = 0 \nonumber
\end{equation}
%
There's no $B$ term, so $-4AC < 0$ since $D$ is positive.

%-------------------------------------------------------------
\vspace*{1em}
\noindent \underline{Elliptic example}:

Poisson's equation (stationary heat conduction with a volumetric source:
%
\begin{equation}
\nabla \cdot (k \nabla T) + q''' = 0 \nonumber
\end{equation}
% 
where
\begin{itemize}
\item $T$ is temperature
\item $k$ is thermal conductivity
\item $q'''$ is volumetric heat generation rate
\end{itemize}
%
recall
%
\begin{align}
\text{gradient is } \nabla T &= \vec{i}\frac{\partial T}{\partial x} + \vec{j}\frac{\partial T}{\partial y} + \vec{k}\frac{\partial T}{\partial z} \nonumber \\
%
\text{divergence is } \nabla \cdot \vec{v} &= \frac{\partial v_x}{\partial x} + \frac{\partial v_y}{\partial y} + \frac{\partial v_z}{\partial z} \nonumber
\end{align}
%
There's no $B$ term, so $-4AC < 0$ since $k$ is positive.

\begin{itemize}
\item Dirichlet BCs if $f(a,b) =$ constant
\item Neumann BCs if 
\begin{equation}
\vec{n} \cdot \nabla f = n_x \frac{\partial f} {\partial x} + n_y \frac{\partial f} {\partial y}  +n_z \frac{\partial f} {\partial z} \nonumber
\end{equation}
\end{itemize}

%-------------------------------------------------------------
\vspace*{1em}
\noindent \underline{Parabolic Examples}:

(1) Transient heat conduction in a slab:
%
\begin{equation}
\frac{\partial T(x,t)}{\partial t} = \frac{\partial}{\partial x} \frac{k}{\rho c} \frac{\partial}{\partial x} T(x,t) + q'''(x,t) \nonumber
\end{equation}
%
where $\rho$ = mass density and $c$ = specific heat capacity

\noindent There aren't $B$ or $C$ terms, so $-4AC = 0$


\vspace*{2em}
(2) Time-dependent, 1-group neutron diffusion equation
%
\begin{equation}
\frac{1}{v} \frac{\partial \phi(x,t)}{\partial t} = \frac{\partial}{\partial x} D \frac{\partial}{\partial x} \phi(x,t) + (\nu \Macro_f - \Macro_a) \phi(x,t) + S(x,t) \nonumber
\end{equation}
%
There aren't $B$ or $C$ terms, so $-4AC = 0$

%-------------------------------------------------------------
\vspace*{1em}
\noindent \underline{Hyperbolic Examples}:

(1) The wave equation
%
\begin{equation}
\frac{\partial^2 f}{\partial t^2} - c \frac{\partial^2 f}{\partial x^2} = 0
\end{equation}
%
where $f(x,t)$ is the displacement and $c$ is the speed of wave propagation.

\noindent There's no $B$ term, and $C$ is negative so $-4AC > 0$

%\noindent \textbf{1st order Hyperbolic equations}
%
%For a system of the form
%%
%\begin{equation}
%\frac{\partial u}{\partial t} + A \frac{\partial u}{\partial x} + B u = F(t,x)
%\end{equation}
%%
%is hyperbolic if the matrix $A$ is diagonalizable with real eigenvalues.
%
%Diagonalizable means there is a nonsingular matrix $P$ such that
%%
% \begin{align}
%    PA^{-1}P &=    &=    \begin{pmatrix}
%      a_{1} & 0 & 0 & \cdots & 0 \\
%      0 & a_{2} & 0 & \cdots & 0 \\
%      0 & 0 & a_{3} & \cdots & 0 \\
%      \vdots & \vdots & \vdots & \ddots   & \vdots \\
%      0 & 0 & 0 & \cdots & a_{d} \\
%    \end{pmatrix} \nonumber  & = \lambda
%    %
%\end{align} 
%http://www.siam.org/books/textbooks/OT88sample.pdf

%\vspace*{2em}
%(2) 1-D, unsteady flow of compressible fluids (in a uniform conduit)
%%
%\begin{equation}
%\frac{\partial \rho}{\partial t} + \gamma \frac{\partial \rho}{\partial x} + \rho \frac{\partial v}{\partial x} = 0
%\end{equation}
%%
%where $\rho$ is the fluid density and $v$ is the fluid velocity.

%\vspace*{2em}
%(2) The equation of motion under the influence of gravity 


%-------------------------------------------------------------
\section{Integro-Differential Equations}

Equations containing both derivatives and integrals. An example we care about is the 1-group, time-independent neutron transport equation in slab geometry
%
\begin{equation}
\mu \frac{\partial \psi(x,\mu)}{\partial x} + \Macro_t \psi(x,\mu) = \frac{\Macro_s}{2} \int_{-1}^1 d\mu' \psi(x, \mu') + S(x, \mu) \nonumber
\end{equation}
%
where $\psi$ is angular neutron flux and $\mu$ is the cosine of the angle

The general energy-dependent, time-dependent, 3-D neutron transport equation is 
\begin{align}
\frac{1}{v} \frac{\partial}{\partial t}\psi(\vec{r}, \hat{\Omega}, E, t) &+ [\hat{\Omega} \cdot \nabla + \Macro(\vec{r}, E)] \psi(\vec{r}, \hat{\Omega}, E, t)  =  \int_0^{\infty} dE' \int_{4\pi} d\hat{\Omega'} \:\Macro_{s}(\vec{r}, E' \to E, \hat{\Omega'} \cdot \hat{\Omega}) \psi(\vec{r}, \hat{\Omega'}, E', t) \nonumber \\
   %
&+\frac{ \chi(E)}{4 \pi} \int_0^{\infty} dE' \:\nu \Macro_{f}(\vec{r}, E') \int_{4\pi} d\hat{\Omega'} \:\psi(\vec{r}, \hat{\Omega'}, E', t) \nonumber
\end{align}
%
\noindent where the quantities are at location $\vec{r}$, travelling in directions $\hat{\Omega}$, at energy $E$ and time $t$ and are defined as:
\begin{list}{}{\hspace{2em}}
  \item $\psi(\vec{r}, \hat{\Omega}, E, t)$ is the angular neutron flux in neutrons per unit length squared per steradian per second and expresses where all the neutrons are in phase space, 
  \item $\chi(E)$ is the fission spectrum and specifies the energy distribution of neutrons born from fission,
%  \item $k$ can be thought of as the asymptotic ratio of the number of neutrons in one generation to the number in the next.
\end{list}

%-------------------------------------------------------------
\section{Integral Equations}

Steady state integral neutron transport equation
%
\begin{equation}
\phi(\vec{r}, E) = \int_V d\vec{r'} \int dE' K(\vec{r}, E, \vec{r'}, E') \phi(\vec{r'}, E') + f(\vec{r}, E) + S(\vec{r}, E) \nonumber
\end{equation}
%
where
%%
\begin{align}
K(\vec{r}, E, \vec{r'}, E') &= \frac{\mathsf{e}^{\tau(\vec{r}, \vec{r'}, E)}}{4\pi |\vec{r'} - \vec{r}|^2} \Macro_s(\vec{r}, E' \rightarrow E) \nonumber \\
%
\tau(\vec{r}, \vec{r'}, E) &= \int_0^R d\vec{r'} \Macro_t(\vec{r'}, E) \qquad \text{``optical length"} \nonumber \\
%
f(\vec{r}, E) &= \int_V d\vec{r'}\frac{\mathsf{e}^{\tau(\vec{r}, \vec{r'}, E)}}{4\pi |\vec{r'} - \vec{r}|^2} F(\vec{r}, E) \qquad \text{fission source}\nonumber \\
%
S(\vec{r}, E) &= \int_S d\vec{r'}\frac{\mathsf{e}^{\tau(\vec{r}, \vec{r'}, E)}}{2\pi |\vec{r'} - \vec{r}|^2} J(\vec{r}, E) \qquad \text{``surface" source, i.e\ incoming current} \nonumber
\end{align}

\end{document}