\NeedsTeXFormat{LaTeX2e}
\documentclass[a4paper, 12 pt]{curve}
%\documentclass[12pt]{article}
%\textwidth=7in
%\textheight=9.5in
%\topmargin=-1in
%\headheight=0in
%\headsep=.5in
%\hoffset=-.85in

%\usepackage[cm]{fullpage}
\usepackage[top=0.75in, bottom=0.75in, left=1in, right=1in]{geometry}
\pagestyle{empty}
\usepackage{tabu}
\parindent 0ex
\usepackage{hyperref}

\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\begin{document}

\begin{center}
{\bf NE 155\\ Analysis Final Project Rubric
}
\end{center}

\setlength{\unitlength}{1in}
\begin{picture}(6,.1) 
\put(0,0) {\line(1,0){6.25}}         
\end{picture}

\renewcommand{\arraystretch}{2}

The final paper should be $\sim$6-7 pages with 1.5 spacing per team member. This may vary based on the specific project; please use your best judgment. Please include these items as clearly labeled sections in the \underline{final report}:
%
\begin{enumerate}
\item \textbf{Introduction:} What are you trying to accomplish and why? Also preview what you are going to talk about.

\item \textbf{Problem Description:} Describe the problem you are solving in this work and explain how it will help you find out the thing you told us about in the introduction.

\item \textbf{Description of Work:} What did you do to perform your analysis? This can include any models you built, data you collected, strategies you needed for evaluation, etc.

\item \textbf{Results:} What did you find out?

\item \textbf{Conclusions:} What do your findings mean? How does that relate to the goal you laid out in the introduction?

\item \textbf{References:} You must have references that you cite in your paper (including the interim report).
\end{enumerate}

\vspace*{1em}
In your \underline{mid-project report}, please replace ``Conclusions" with \textbf{Plans for Completion} and keep in mind that ``Results" will be preliminary or possibly empty. The first three sections don't have to be completely polished, but they should at least be very solid drafts. I will provide feedback, so the better they are when I read them the more useful the feedback will be. This should be a maximum of 4, 6, or 8 pages of 1.5 spacing for 1, 2, or 3 people, respectively.

\vspace*{2em}
Please include these same items in your \underline{final presentation}. The presentations should be approximately 8, 12, or 15 minutes for 1, 2, or 3 people, respectively. This may vary based on the specific project; please use your best judgment.

\vspace*{2em}
\input{notes}

\vspace*{2em}
I will use the following rubric for evaluating the paper:

\vspace*{1 em}
\begin{center}
\begin{tabu}{| X | c | c |}\hline
\textbf{Category} & \textbf{Possible Points} & \textbf{Earned Points} \\ \hline \hline
Work implemented correctly & 15 & \\ \hline
Goal, problem solved, and analysis conducted have an appropriate logical flow & 12 & \\ \hline
Conclusions are supported by the results & 10 & \\ \hline
Complete sentences; correct grammar and spelling & 8 & \\ \hline
Sources properly documented & 5 & \\ \hline
Total & 50 & \\\hline
\end{tabu} 
\end{center}

This rubric is for evaluating the presentation:

\vspace*{1 em}
\begin{center}
\begin{tabu}{| X | c | c |}\hline
\textbf{Category} & \textbf{Possible Points} & \textbf{Earned Points} \\ \hline \hline
Topic motivation is clear & 3 & \\ \hline
Explanation of work is understandable, correct, and supports the motivation & 7 & \\ \hline
Results and conclusions are clearly communicated & 6 & \\ \hline
Good presentation skills: eye contact, volume, clarity of slides, etc. & 6 & \\ \hline
Appropriate presentation length & 3 & \\ \hline
Total & 25 & \\\hline
\end{tabu} 
\end{center}


\end{document}